Analysis of protein tunnel traversability for flexible ligands using motion planning

### 2017-05-05 paper.tex
intro

\title{
    Analysis of protein tunnel traversability for flexible ligands
}

@@

\begin{abstract}
Proteins are involved in many biochemical processes.
%Behavior and properties of proteins as well as other bio-macromolecules are influenced by internal void space such as tunnels or cavities.
Behavior of proteins is influenced by internal void space such as tunnels or cavities.
Tunnels are paths leading from an active site inside the protein to its surface.
Knowledge about tunnels and their evolution in time provides an insight into protein properties
(e.g., stability).
Sampling-based motion planning methods like Rapidly Exploring Random Tree (RRT) can be used to find the tunnels by
sampling the corresponding configuration space.
The main advantage of these methods is that they can be easily adapted to consider protein dynamics.
%i.e., they can find tunnels in proteins moving in time.
However, the inner void space of proteins is  very limited, 
    which may decrease ability of RRT to find a solution due to the narrow passage problem.
In this paper, we propose to generate the random samples using Voronoi Diagram of the atoms.
A subset of Voronoi vertices is dynamically maintained to support generation of samples in promising regions inside the protein.

\end{abstract}

@@

%For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
%This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
%The active site is a specific site, usually deeply buried inside the protein, where the chemical reaction between the protein and ligand can undergo.
%Since the importance of the presence of tunnels in proteins was revealed, the researchers started to study them intensively.
%Koudel\'{a}kov\'{a} et al.~\cite{Koudelakova2013} show that mutations of amino acids located directly around the tunnel improved the structural and kinetic stability of the studied protein substantially, while the surface mutations almost did not contribute to protein stabilization.
%The importance of tunnel can be demonstrated on an example, 
% where mutations of amino acids located directly around the tunnel improved the structural 
%and kinetic stability of the studied protein substantially, while surface mutations almost did not contribute to
%the protein stabilization~\cite{Koudelakova2013}.

%On the other hand, surface mutations almost did not contribute to protein stabilization~\cite{Koudelakova2013}.

%The task of tunnel detection is to find a collision-free path leading from an active site inside the protein to its surface.
%As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
%the radius of the smallest sphere that can traverse the whole tunnel)

Generally, the task of tunnel detection is to find a collision-free path leading from the active site to the protein's surface. 
The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
An example is depicted in Fig.~\ref{fig::motiv}.
Early solutions focused on static molecules. 
However, from a static snapshot it is hard to assess the biochemical relevance of the tunnel as it does not
provide information about its temporal stability.

@@

%The ability of RRT to quickly explore the configuration space is caused by the combination
%of uniform sampling and the nearest-neighbor rule used to select nodes for the expansion.
The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
A narrow passage is a small region of the configuration space containing part of the solution.
As the samples are generated from uniform distribution in RRT, probability of placing the samples into the narrow passages is low and therefore,
   the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
%Due to its low volume, the probability of expanding the nodes towards the passage is small~\cite{hannaWIS}.
%    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.
A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
%This requires to precompute the medial axis, which can be time consuming.
%Another approach is to shift the random samples towards the medial axis~\cite{amatoOBPRM} or into a direction of estimated
%medial axis~\cite{hollemanMAPRM}.
Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
Increased probability of sampling in a given region, e.g., in a narrow passage, 
brings advantage only if the tree is close to the region and if it can reach it.

@@

>>>>

\title{
    Analysis of protein tunnel traversability for flexible ligands using motion planning
}

@@

\begin{abstract}
%Proteins are involved in many biochemical processes.
Behavior and properties of proteins as well as other bio-macromolecules are influenced by internal void space such as tunnels or cavities.
%Behavior of proteins is influenced by internal void space such as tunnels or cavities.
Tunnels are paths leading from an active site inside the protein to its surface and they act as transporting pathways
to the active site.
To achieve a desired propery of the proteins, the active sites has to be reached by small molecules (ligands).
The widely used software tools for tunnel detection are based on Voronoi diagrams, which is suitable
for detecting tunnel of spherical probes.
It is however not easy to estimate if a given non-spherical ligand can pass a spherical tunnel.
Moreover, ligands are not rigid structures, but their conformation changes due to interactions with the protein's atoms.
In this paper, we propose to employ sampling-based motion planning techniques to analyze traversability of ligands in the spherical
tunnels.
To speed up the computation, the flexibility of ligands is modeled using a predefined set of conformations suitable (typical) for 
the given ligand.
We introduce a novel modification of RRT method, whose search in the corresponding configuration space is guided by
the tunnel pathway.
\end{abstract}

@@

For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
The active site is a specific site, usually deeply buried inside the protein, where the chemical reaction between the protein and ligand can undergo.
Since the importance of the presence of tunnels in proteins was revealed, the researchers started to study them intensively.
%Koudel\'{a}kov\'{a} et al.~\cite{Koudelakova2013} show that mutations of amino acids located directly around the tunnel improved the structural and kinetic stability of the studied protein substantially, while the surface mutations almost did not contribute to protein stabilization.
%The importance of tunnel can be demonstrated on an example, 
% where mutations of amino acids located directly around the tunnel improved the structural 
%and kinetic stability of the studied protein substantially, while surface mutations almost did not contribute to
%the protein stabilization~\cite{Koudelakova2013}.

The task of tunnel detection is to find a collision-free path leading from an active site inside the protein to its surface.
An example is depicted in Fig.~\ref{fig::motiv}.
As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
the radius of the smallest sphere that can traverse the whole tunnel)

Available tools for tunnel detection compute the tunnels assuming spherical ligands (probes). 
Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel
lengths and its bottlenect.
Such decision is however not precise and require previous expertise with the domain.

To provide more information about the traversability of the tunnels, it is necessary to consider a non-spherical model of 
the ligand.
Such task can be then solved using motion planning techniques.
It is necessary to consider translation, rotation and possibly also other degrees of freedom of the ligands, that are determined
by the torsional angles, the related configuration space, that need to be searched, is high-dimensional. 
Sampling-based motion planning techniques can be used to find a trajectory of the ligand.

Generally, the task of tunnel detection is to find a collision-free path leading from the active site to the protein's surface. 
The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
Early solutions focused on static molecules. 
However, from a static snapshot it is hard to assess the biochemical relevance of the tunnel as it does not
provide information about its temporal stability.

@@

%The ability of RRT to quickly explore the configuration space is caused by the combination
%of uniform sampling and the nearest-neighbor rule used to select nodes for the expansion.
The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
A narrow passage is a small region of the configuration space containing part of the solution.
As the samples are generated from uniform distribution in RRT, probability of placing the samples into the narrow passages is low and therefore,
   the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
%Due to its low volume, the probability of expanding the nodes towards the passage is small~\cite{hannaWIS}.
%    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.

A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
%This requires to precompute the medial axis, which can be time consuming.
%Another approach is to shift the random samples towards the medial axis~\cite{amatoOBPRM} or into a direction of estimated
%medial axis~\cite{hollemanMAPRM}.
Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
Increased probability of sampling in a given region, e.g., in a narrow passage, 
brings advantage only if the tree is close to the region and if it can reach it.
To utilize a knowledge about the environment in the sampling process, represented e.g. using a mexial axis or a 2D or 3D path,
it is necessary to maintain the sampling regions according the progress of the tree.
For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
the path.
Similarly in
\cite{jorry14marrt,neddy2016dynnamic}
ob-rrt
rrt-retraction

@@

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::tunnel}expand()}
\KwIn{
   configuration $q$ to be expanded
}
%\KwData{
%}
\KwOut{
    set $S$ of collision-free configurations reachable from $q$
}
\hrule
$c = $ conformation of $q$\;
\ForEach{$c_j \in C$}{
    \If{$canSwtich(c,c_j)$}{
        \ForEach{$s \in scales$}{
            \For{i = 1:m}{
                $q' = q$\;
                translate by random around $q'$\;
                rotate randomly \;
                $q'.c = c_j$\;
                $q'.s = s$\;
                \If{$isCollisionFree(q')$}{
                    $S = S \cup \{ q \}$\;
                }
            }
        }
    }
}
\end{algorithm}
